Time stretching of the GeV emission of GRBs: Fermi-LAT data vs geometrical model

### 2015-11-30 Paper/gammaRays.tex
Add references from the 2nd physics reviewer

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\cite{Vianello:2013ela,Gehrels:2013xd} for a
review. In particular, the total energy emitted in gamma-rays during a
burst was found to be similar for the different GRBs within an order
of magnitude~\cite{Bloom:2003wy}. This indicates that the most of the
bursts have similar energetics in their rest frames. Moreover, the
high energy radiation of GRBs is shown to follow the predictions of
the synchrotron radiation scenario~\cite{Wang:2013ptaw}.  Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands
\cite{Yi:2005ht} and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\cite{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this paper
we go further and compare the temporal extension of GRB gamma-ray
radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\cite{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors \cite{Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba}.  These
studies concerned x-ray radiation, and considered the homogeneous
distribution of radiation sources throughout the jet. Our model is
based on the reverse assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum.

@@

We still have one ingredient missing though -- the evolution of the
bursts density. We assume that the bursts density is roughly
proportional to the stars density, and the latter is roughly
proportional to the matter density, which changes with $z$ as
$\left(1+z\right)^3$. It is clear, however, that since no stars
existed at very small redshifts, the burst density should decline
there, so we add an exponential cutoff to it:
\begin{equation}\label{eq:redshift}
\rho = \rho_0 \left(1+z\right)^3 \exp\left(-\frac{z}{z_c}\right)\,
\end{equation}
where $\rho_0$ is a normalization factor, and $z_c$ is a redshift
scale, after which the density is cut off.

@@

First of all, we need to determine, which parameters to fit. At most,
the observables of a particular burst depend on 10 parameters: 8 of
the burst $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k,
\alpha\right)$, and 2 of the observer's position related to the burst
$\left(z, \chi\right)$. However, some of these parameters only change
observables trivially, and, also, there are transformations of
variables, which do not change observables at all (see equation
\ref{eq:p}):
\begin{itemize}
	\item{The number of photons $p$ depends linearly on $\eta_0$,
          so $\eta_0$ doesn't affect other observables like duration,
          or stretching factor. One can easily fit $\eta_0$ to match
          the observed photon count.}
	\item{If we make a simultaneous transformation of $r_0
          \rightarrow \lambda_1 r_0$ and $\eta_0 \rightarrow
          \frac{1}{\lambda_1^3}\eta_0$, where $\lambda_1$ is a parameter
          of transformation, only duration of the burst will
          change. Other observables, like the total number of photons,
          or stretching factor, will not be affected.}
	\item{Finally, if we make a following transformation:
          $\omega_0 \rightarrow \lambda_2 \omega_0$, $\theta_0
          \rightarrow \lambda_2^k \theta_0$, $\eta_0 \rightarrow
          \lambda_2^\alpha \eta_0$, where $\lambda_2$ is another
          transformation parameter, no observables will change
          whatsoever.}
\end{itemize}
Given these transformations, one can reduce the number of parameters
for fit to 7: $\left(\gamma, n, \theta_0, k, \alpha, z, \chi\right)$.

@@

We are going to fit parameters specifically for GRB 090926A for its stretching factor is the most significant. While we assume burst related parameters $\left(n, \theta_0, k, \alpha\right)$ to be the same for all bursts, $\chi$ here considered to be the off-axis angle of GRB 090926A. At the same time, we need to make sure that burst related universal parameters $\left(n, \theta_0, k, \alpha\right)$ are consistent with observation of other bursts such as GRB 090902B.

In particular, the best fit parameters are found by minimizing the cost function
$C\left(n, \theta_0, k, \alpha, \chi\right)$ which
satisfies the following objectives:
\begin{itemize}
	\item { Total energy of the burst is finite (that is
          $k+\frac{\alpha}{2}+1 < 0$) and smaller than $\unit[6 \times
            10^{53}]{GeV} \approx \unit[10^{51}]{erg}$
          \cite{Gehrels:2013xd}.  }
	\item { The stretching factor of the burst should be
          compatible with the value for the GRB 090926A.  }
	\item { The ratio of photons with high energies and low
          energies should be compatible with the value for the GRB
          090926A, $\frac{p_{\infty}
            \left(z,\chi;\unit[1]{GeV},\infty\right)}{p_{\infty}
            \left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV}\right)} =
          0.057$.  }
	\item { The burst should not be too faint compared to
          other bursts from the sample. In other words, the total
          number of observed photons should approximately equal to the
          median among the sample.  }
	\item { We assume that all bursts have the same burst
          parameters $\left(\gamma, \eta_0, r_0, n, \omega_0,
          \theta_0, k, \alpha\right)$ and require that bursts with
          small stretching factors like GRB 090902B should be possible
          to observe. So we require that bursts with the stretching
          factor of GRB 090902B or lower appear in random samples by
          varying $z$ and $\chi$ (sampling is done the same way as in
          Section \ref{sec:distribution}). Note that this requirement
          unlike others applies to an ensemble of simulated bursts. }
\end{itemize}

>>>>

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\cite{Vianello:2013ela,Gehrels:2013xd} for a
review. In particular, the total energy emitted in gamma-rays during a
burst was found to be similar for the different GRBs within an order
of magnitude~\cite{Bloom:2003wy}. This indicates that the most of the
bursts have similar energetics in their rest frames. {\bf Also, it is generally accepted that} the
high energy radiation of GRBs is shown to follow the predictions of
the synchrotron radiation scenario~\cite{Wang:2013ptaw}, {\bf although there are problems with that model~\cite{Daigne:2010fb} and there are bursts in which additional radiation component clearly appears~\cite{Abdo:2009pg} }.  Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands {\bf 
\cite{cheng1995time,Wu:1999ms} }and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\cite{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}. In this paper
we go further and compare the temporal extension of GRB gamma-ray
radiation in the two energy bands $\unit[100]{MeV} < E <
\unit[1]{GeV}$ and $E > \unit[1]{GeV}$ (hereafter, low and high-energy
bands) using the data from the Large Area Telescope (LAT) of the Fermi
satellite~\cite{2009ApJ...697.1071A,Ackermann:2012kna}.

@@

We propose that the observed energy dependence of the time profiles
may be explained without introduction of the new effects. We argue
that the stretching may originate from the effects of the jet geometry
called curvature effects. These effects were explored by multiple
authors{\bf \cite{Meszaros:1997je,Soderberg:2001mk,Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba} }.  These
studies concerned x-ray radiation, and considered the homogeneous
distribution of radiation sources throughout the jet. Our model is
based on the reverse assumption that the highest energy radiators are
concentrated near the axis of the jet and consequently the jet opening
angle depends on energy. The latter is motivated by observations if
all bursts are considered equivalent. In this case the energy
dependence of the jet opening angle resolves the contradiction between
the small fraction of GRBs visible above GeV and the hard
spectrum.

@@

We still have one ingredient missing though -- the evolution of the
bursts density. We assume that the bursts density is roughly
proportional to the stars density, and the latter is roughly
proportional to the matter density, which changes with $z$ as
$\left(1+z\right)^3$. It is clear, however, that since no stars
existed at very {\bf large} redshifts, the burst density should decline
there, so we add an exponential cutoff to it:
\begin{equation}\label{eq:redshift}
\rho = \rho_0 \left(1+z\right)^3 \exp\left(-\frac{z}{z_c}\right)\,
\end{equation}
where $\rho_0$ is a normalization factor, and $z_c$ is a redshift
scale, after which the density is cut off.

@@

First of all, we need to determine, which parameters to fit. At most,
the observables of a particular burst depend on 10 parameters: 8 of
the burst $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k,
\alpha\right)$, and 2 of the observer's position related to the burst
$\left(z, \chi\right)$. However, some of these parameters only change
observables trivially, and, also, there are transformations of
variables, which do not change observables at all (see equation
\ref{eq:p}):
\begin{itemize}
	\item{The number of photons $p$ depends linearly on $\eta_0$,
          so $\eta_0$ doesn't affect other observables like duration,
          or stretching factor. One can easily fit $\eta_0$ to match
          the observed photon count.}
	\item{{\bf If we make a simultaneous transformation of $r_0
          \rightarrow \lambda_1 r_0$ and $\eta_0 \rightarrow
          \frac{1}{\lambda_1^3}\eta_0$, where $\lambda_1$} is a parameter
          of transformation, only duration of the burst will
          change. Other observables, like the total number of photons,
          or stretching factor, will not be affected.}
	\item{{\bf Finally, if we make a following transformation:
          $\omega_0 \rightarrow \lambda_2 \omega_0$, $\theta_0
          \rightarrow \lambda_2^k \theta_0$, $\eta_0 \rightarrow
          \lambda_2^\alpha \eta_0$, where $\lambda_2$ } is another
          transformation parameter, no observables will change
          whatsoever.}
\end{itemize}
Given these transformations, one can reduce the number of parameters
for fit to 7: $\left(\gamma, n, \theta_0, k, \alpha, z, \chi\right)$.

@@

{\bf We are going to fit parameters specifically for GRB 090926A for its stretching factor is the most significant. While we assume burst related parameters $\left(n, \theta_0, k, \alpha\right)$ to be the same for all bursts, $\chi$ here considered to be the off-axis angle of GRB 090926A. At the same time, we need to make sure that burst related universal parameters $\left(n, \theta_0, k, \alpha\right)$ are consistent with observation of other bursts such as GRB 090902B.

In particular, the best fit parameters are found by minimizing the cost function
$C\left(n, \theta_0, k, \alpha, \chi\right)$ which
satisfies the following objectives:}
\begin{itemize}
	\item { Total energy of the burst is finite (that is
          $k+\frac{\alpha}{2}+1 < 0$) and smaller than $\unit[6 \times
            10^{53}]{GeV} \approx \unit[10^{51}]{erg}$
          \cite{Gehrels:2013xd}.  }
	\item { The stretching factor of the burst should be
          compatible with the value for the GRB 090926A.  }
	\item { The ratio of photons with high energies and low
          energies should be compatible with the value for the GRB
          090926A, $\frac{p_{\infty}
            \left(z,\chi;\unit[1]{GeV},\infty\right)}{p_{\infty}
            \left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV}\right)} =
          0.057$.  }
	\item { The burst should not be too faint compared to
          other bursts from the sample. In other words, the total
          number of observed photons should approximately equal to the
          median among the sample.  }
	\item { We assume that all bursts have the same burst
          parameters $\left(\gamma, \eta_0, r_0, n, \omega_0,
          \theta_0, k, \alpha\right)$ and require that bursts with
          small stretching factors like GRB 090902B should be possible
          to observe. So we require that bursts with the stretching
          factor of GRB 090902B or lower appear in random samples by
          varying $z$ and $\chi$ (sampling is done the same way as in
          Section \ref{sec:distribution}). Note that this requirement
          unlike others applies to an ensemble of simulated bursts. }
\end{itemize}
